---
title: "Getting Started with ggpatchy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ggpatchy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
```

ggpatchy adds a `pattern` aesthetic to ggplot2 geoms. Patterns are rendered
via native grid graphics — no ImageMagick or external raster dependencies are
required.

```{r load}
library(ggplot2)
library(ggpatchy)
```

## Bar and column charts

`geom_col_pattern()` and `geom_bar_pattern()` are drop-in replacements for
`geom_col()` and `geom_bar()`. Map `pattern` to a discrete variable with
`scale_pattern_manual()` or `scale_pattern_discrete()`.

```{r bar}
df <- data.frame(
  category = c("A", "B", "C", "D"),
  value    = c(3.2, 5.1, 4.0, 2.7)
)

ggplot(df, aes(category, value, fill = category, pattern = category)) +
  geom_col_pattern() +
  scale_pattern_manual(values = c(
    A = "hatch", B = "crosshatch", C = "dots", D = "vertical"
  )) +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal() +
  labs(title = "geom_col_pattern()")
```

`geom_bar_pattern()` uses `stat = "count"`:

```{r bar-count}
ggplot(mpg, aes(class, fill = class, pattern = class)) +
  geom_bar_pattern() +
  scale_pattern_discrete() +
  scale_fill_brewer(palette = "Set3") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(title = "geom_bar_pattern() with scale_pattern_discrete()")
```

## Polygons

`geom_polygon_pattern()` clips patterns to the exact polygon boundary, not
just the bounding box. This means concave shapes like crescents or L-shapes
are handled correctly.

```{r polygon}
theta <- seq(0, 2 * pi, length.out = 7)[-7]
hex <- data.frame(x = cos(theta), y = sin(theta))

ggplot(hex, aes(x, y)) +
  geom_polygon_pattern(
    pattern        = "hatch",
    fill           = "lightyellow",
    pattern_colour = "goldenrod",
    pattern_angle  = 30,
    pattern_spacing = 0.07
  ) +
  coord_fixed() +
  theme_void() +
  labs(title = "geom_polygon_pattern() on a hexagon")
```

## Ribbon charts

`geom_ribbon_pattern()` adds patterns to confidence bands or any region
defined by `ymin` and `ymax`.

```{r ribbon}
x  <- seq(0, 2 * pi, length.out = 80)
df <- data.frame(x = x, ymin = sin(x) - 0.3, ymax = sin(x) + 0.3)

ggplot(df, aes(x, ymin = ymin, ymax = ymax)) +
  geom_ribbon_pattern(
    pattern        = "hatch",
    fill           = "lightblue",
    pattern_colour = "steelblue"
  ) +
  theme_minimal() +
  labs(title = "geom_ribbon_pattern() — confidence band")
```

Multiple groups with different patterns:

```{r ribbon-groups}
df2 <- data.frame(
  x       = c(x, x),
  ymin    = c(sin(x) - 0.2, cos(x) + 0.5),
  ymax    = c(sin(x) + 0.2, cos(x) + 1.0),
  group   = rep(c("sin", "cos"), each = length(x)),
  pattern = rep(c("hatch", "crosshatch"), each = length(x)),
  fill    = rep(c("lightblue", "lightyellow"), each = length(x))
)

ggplot(df2, aes(x, ymin = ymin, ymax = ymax,
                group = group, fill = fill, pattern = pattern)) +
  geom_ribbon_pattern(alpha = 0.8) +
  scale_pattern_identity(guide = "legend") +
  scale_fill_identity() +
  theme_minimal() +
  labs(title = "geom_ribbon_pattern() — two groups")
```

## Area charts

`geom_area_pattern()` mirrors `geom_area()`: it sets `ymin = 0` automatically
and defaults to `position = "stack"`.

```{r area}
df3 <- data.frame(
  x       = rep(1:8, 2),
  y       = c(1, 3, 2, 4, 3, 2, 4, 3,
              2, 1, 3, 2, 1, 3, 2, 1),
  group   = rep(c("Group A", "Group B"), each = 8),
  pattern = rep(c("hatch", "crosshatch"), each = 8)
)

ggplot(df3, aes(x, y, fill = group, pattern = pattern)) +
  geom_area_pattern(position = "stack", alpha = 0.85) +
  scale_pattern_identity(guide = "legend") +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal() +
  labs(title = "geom_area_pattern() — stacked")
```

## Violin charts

`geom_violin_pattern()` adds patterns to violin plots. Use it as a drop-in
replacement for `geom_violin()`.

```{r violin}
ggplot(mpg, aes(class, hwy, fill = class, pattern = class)) +
  geom_violin_pattern(pattern_colour = "grey30") +
  scale_pattern_discrete() +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(title = "geom_violin_pattern()")
```

## Density plots

`geom_density_pattern()` extends `geom_density()` for pattern fills on smooth
density estimates.

```{r density}
ggplot(mpg, aes(hwy, fill = drv, pattern = drv)) +
  geom_density_pattern(alpha = 0.7, pattern_colour = "grey20") +
  scale_pattern_manual(values = c("4" = "hatch", "f" = "crosshatch", "r" = "dots")) +
  scale_fill_brewer(palette = "Pastel2") +
  theme_minimal() +
  labs(title = "geom_density_pattern()")
```

## Rectangles and tiles

`geom_rect_pattern()` and `geom_tile_pattern()` add pattern overlays to
rectangle geoms. `geom_tile_pattern()` is convenient for heatmap-style layouts
where cells are specified by centre coordinates.

```{r rect}
blocks <- data.frame(
  xmin    = c(0, 1, 2, 0, 1, 2),
  xmax    = c(1, 2, 3, 1, 2, 3),
  ymin    = c(0, 0, 0, 1, 1, 1),
  ymax    = c(1, 1, 1, 2, 2, 2),
  pattern = c("hatch", "crosshatch", "dots", "vertical", "horizontal", "weave")
)

ggplot(blocks, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax,
                   fill = pattern, pattern = pattern)) +
  geom_rect_pattern(pattern_colour = "grey30") +
  scale_pattern_identity() +
  scale_fill_brewer(palette = "Pastel1") +
  coord_fixed() +
  theme_void() +
  labs(title = "geom_rect_pattern() — all six line/dot patterns")
```

## Choropleth maps

`geom_sf_pattern()` works with sf geometry columns. It uses pattern to add a
second encoding channel alongside fill colour, which is useful for
accessibility (grayscale printing, colour-vision deficiency). See the
**Mapping with Patterns** vignette for full examples with the NC counties and
US states datasets.

## Pattern reference

Built-in patterns:

| Name         | Description                         | Respects `pattern_angle`? |
|:-------------|:------------------------------------|:--------------------------|
| `none`       | No pattern (fill only)              | —                         |
| `hatch`      | Diagonal lines                      | Yes                       |
| `crosshatch` | Crossed diagonal lines              | Yes                       |
| `horizontal` | Horizontal lines                    | No (fixed)                |
| `vertical`   | Vertical lines                      | No (fixed)                |
| `dots`       | Regular dot grid                    | No (fixed)                |
| `weave`      | Woven over/under line texture       | No (fixed)                |

`pattern_spacing` is interpreted as a fraction of each shape's bounding box,
so the visual density stays consistent across shapes of any size. Custom
patterns can be registered with `register_pattern()`.

## Known limitations

**`stat = "identity"` with `geom_violin_pattern()` requires a `violinwidth`
column.** The default stat (`"ydensity"`) computes `violinwidth` automatically
from your raw data. If you pass `stat = "identity"` with pre-summarised density
data, your data frame must include a `violinwidth` column (values in [0, 1]
giving the normalised half-width at each `y` level). If the column is absent,
`geom_violin_pattern()` emits an informative message and skips that group
rather than crashing.

**`orientation = "y"` (horizontal) is not supported for pattern overlays.**
`geom_violin_pattern()`, `geom_ribbon_pattern()`, and `geom_area_pattern()`
all skip the pattern when `orientation = "y"` or `flipped_aes = TRUE`. The
base shape still renders correctly. Use `coord_flip()` on a vertical geom as
a workaround.

**Non-Cartesian coordinates:** Patterns are drawn in screen (npc) space after
`coord$transform()`. Under `coord_polar` or `coord_sf` the pattern lines
remain straight in screen pixels rather than following the coordinate
curvature.
