---
title: "Mapping with Patterns"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mapping with Patterns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%",
  dpi       = 150
)

has_sf     <- requireNamespace("sf",     quietly = TRUE)
has_spdata <- requireNamespace("spData", quietly = TRUE)
```

Pattern fill overlays give choropleth maps a second encoding channel that
works independently of colour. The result reads correctly in grayscale, in
PDF, and for readers with colour-vision deficiency — three contexts where a
colour-only map fails.

## North Carolina counties

The North Carolina county dataset bundled with the **sf** package lets us
encode two census variables simultaneously: county area (`AREA`, continuous)
mapped to fill colour, and 1974 birth-rate tertile (`BIR74`, cut into three
equal-count bins) mapped to pattern. Readers can see both dimensions at
once — the pattern distinguishes high-birth-rate counties regardless of
whether they also happen to be large or small.

```{r nc-map, eval = has_sf}
library(ggplot2)
library(sf)
library(ggpatchy)

nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)

# Quantile-based tertiles give balanced bin sizes across skewed BIR74 data
nc$birth_bin <- cut(
  nc$BIR74,
  breaks        = quantile(nc$BIR74, probs = c(0, 1/3, 2/3, 1)),
  include.lowest = TRUE,
  labels        = c("Low", "Medium", "High")
)

ggplot(nc) +
  geom_sf_pattern(
    aes(fill = AREA, pattern = birth_bin),
    pattern_colour    = "grey30",
    pattern_linewidth = 0.4,
    pattern_spacing   = 0.12
  ) +
  scale_pattern_manual(
    values = c(Low = "hatch", Medium = "crosshatch", High = "dots"),
    name   = "Birth rate (1974)"
  ) +
  scale_fill_distiller(palette = "Oranges", name = "Area") +
  labs(
    title    = "North Carolina counties",
    subtitle = "Fill: county area  ·  Pattern: 1974 birth-rate tertile"
  ) +
  theme_minimal() +
  theme(axis.text = element_blank())
```

## US states

The **spData** package's `us_states` dataset maps 2015 population
(`total_pop_15`) to fill colour and US Census region to pattern. Four
patterns — one per region — distinguish Midwest, Northeast, South, and West
without adding a second colour scale. The combination lets readers ask two
questions at once: *how populous is this state?* (colour) and *which region
is it in?* (pattern). Both questions are answerable from a black-and-white
printout.

```{r us-map, eval = has_sf && has_spdata}
library(spData)

# Fix a known typo in the REGION factor level
us <- us_states
levels(us$REGION)[levels(us$REGION) == "Norteast"] <- "Northeast"

ggplot(us) +
  geom_sf_pattern(
    aes(fill = total_pop_15, pattern = REGION),
    pattern_colour    = "grey20",
    pattern_linewidth = 0.45,
    pattern_spacing   = 0.09
  ) +
  scale_pattern_manual(
    values = c(
      Midwest   = "hatch",
      Northeast = "crosshatch",
      South     = "dots",
      West      = "vertical"
    ),
    name = "Region"
  ) +
  scale_fill_distiller(
    palette   = "Blues",
    direction = 1,
    name      = "Population (2015)",
    labels    = scales::comma
  ) +
  labs(
    title    = "US states",
    subtitle = "Fill: 2015 population  ·  Pattern: Census region"
  ) +
  theme_minimal()
```
